---
title: "Layman Problem Set 3 - Statistics"
author: Maggie Layman
date: Nov/22/2024
format: html
editor: visual
---

# Problem Set 3

## Here I will demonstrate everything I have learned about basic statistical analysis using R

Workflow model to follow: plot data (and guess the results) → build the statistical model → Check the model assumptions → Interpret the model → Make a final plot

In this project, I will examine data featuring 5 species of bats and use some data analysis to learn more.

start with a clean environment and package installation

```{r}
rm (list = ls()) # cleans environment completely
library(tidyverse)
library(here)
library(ggfortify)
library(multcomp)
```

Next, I'll read in and check out the data I need using the here() function

```{r}
bats<- read.csv(here("data","bat_data.csv"))
glimpse(bats)
```

Taking a quick look at this data - there are 9 variables and 125 rows of data. There are 5 categorical variables and 4 continuous variables.

I want to change at least a few of these data types using as.factor()

```{r}
bats$Species<-as.factor(bats$Species)
bats$common_name<-as.factor(bats$common_name)
bats$age<-as.factor(bats$age)
bats$sex<-as.factor(bats$sex)
bats$type<-as.factor(bats$type)
```

Next, I'll check the factor data for consistency

```{r}
levels(bats$Species)
levels(bats$common_name)
levels(bats$age)
levels(bats$sex)
levels(bats$type)
```

These are clean- no mistakes in entry!

Now to check out the numeric variables - I want to ensure there are no odd outliers made by mistake

```{r}
ggplot(bats,aes(x=mass_g,y=tick_num))+
  geom_point()+
  theme_bw()
```

There don't appear to be any strange outliers here

```{r}
ggplot(bats,aes(x=body_length_mm,y=tail_length_mm))+
  geom_point()+
  theme_bw()
```

Again, nothing concerning in terms of outliers in the data

## Now that I'm confident in the data, I'll start looking into Aria's questions:

Workflow model to follow: plot data (and guess the results) → build the statistical model → Check the model assumptions → Interpret the model → Make a final plot

## **1. Does body mass differ between these five species of bats and, if so, how does body mass differ between the species?**

To figure this out, I'll first plot this proposed relationship, knowing species (the predictor) is a categorical variable and mass (the response) is a continuous variable, so I'll use a boxplot, and will later run an ANOVA (there are 5 treatments of Species).

```{r}
ggplot(bats,aes(x=Species,y=mass_g))+
  geom_boxplot()+ #creates a boxplot
  geom_point()+
  theme_bw()
```

Interesting! I would definitely expect a difference in the mean mass between at least two of these species. At minimum there should be a difference between LACI and LANO, MYLU, and PESU, and at maximum I could see a statistically significant difference between each of them.

Next, I'll build my statistical model. Since I'm going to run an ANOVA, I'll need to start with a linear model

```{r}
bats.MOD <- lm(mass_g~Species, data=bats)
```

I'll need to check my assumptions before running any tests

```{r}
autoplot(bats.MOD)
```

These look great! A linear model is appropriate, there is normality in the data, there is even variance, and no points with major leverage.

Now I can move onto my tests

```{r}
anova(bats.MOD)
```

Woooop! We have a (very) significant p value.

I reject the null hypothesis that mean body mass does not differ between these five species of bats, and suggest that at least one of the species correlates with a different mean body mass (DF120,4, F=792.9, p\<2.2e-16).

Okay, great! But how does body mass differ between the species?

I'll have to run a post-hoc test to find out.

This is a Tukey test

```{r}
PHC1<-glht(bats.MOD, linfct = mcp(Species="Tukey")) #mcp = multiple comparisons
summary(PHC1)

```

I could do an interpretation here, but it is a lot of information to take in, so I'll ask for a compact letter display

```{r}
cld(PHC1)
```

Cool! This tells us that each of the species are statistically different in means of mass.

Let's make a final model to wrap up

```{r}
sumbats1<-bats%>%
  group_by(Species)%>%
  summarise(
    meanbats = mean(mass_g),
    mass_g=mass_g,
    sebats=sd(mass_g)/sqrt(n())
  )

ggplot(sumbats1,aes(x=Species,y=mass_g,colour=Species))+
  geom_point()+
  geom_errorbar(aes(ymin=meanbats-sebats,ymax=meanbats+sebats))+
  theme_bw()
```

## **2. Does body length differ between these five species of bats and, if so, how does body mass differ between the species?**

I'm pretty sure this was supposed to say body **length** twice

To figure this out, I'll first plot this proposed relationship, knowing that species (the predictor) is a categorical variable and body length (the response) is a continuous variable, so I'll use a boxplot, and will later run an ANOVA (there are 5 treatments of Species)

```{r}
ggplot(bats,aes(x=Species,y=body_length_mm))+
  geom_boxplot()+ #creates a boxplot
  geom_point()+
  theme_bw()
```

Reviewing this graph, I would expect a difference in the mean mass between at least two of these species, likely between LACI compared to LANO, MYLU, and PESU. In fact, I would expect difference between them all, with a possible exemption for MYLU and PESU

Next, I'll build my statistical model. Since I'm going to run an ANOVA, I'll need to start with a linear model

```{r}
bats.MOD2 <- lm(body_length_mm~Species, data=bats)
```

I'll need to check my assumptions before running any tests

```{r}
autoplot(bats.MOD2)
```

These look great! A linear model is appropriate, there is normality in the data, there is even variance, and no points with major leverage.

Now I can move onto my tests

```{r}
anova(bats.MOD2)
```

Woooop! We have a (very) significant p value.

I reject the null hypothesis that mean body mass does not differ between these five species of bats, and suggest that at least one of the species correlates with a different mean body mass (DF120,4, F=70.498, p\<2.2e-16).

Okay, great! But how does body length differ between the species?

I'll have to run a post-hoc test to find out.

This is a Tukey test

```{r}
PHC2<-glht(bats.MOD2, linfct = mcp(Species="Tukey")) #mcp = multiple comparisons
summary(PHC2)

```

I could do an interpretation here, but it is a lot of information to take in, so I'll ask for a compact letter display

```{r}
cld(PHC2)
```

Ahhh, this is what I suspected. There is a statistically signficant difference in body length between each of the species other than MYLU and PESU!

New model time!

```{r}
sumbats2<-bats%>%
  group_by(Species)%>%
  summarise(
    meanbats = mean(body_length_mm),
    body_length_mm = body_length_mm,
    sebats=sd(body_length_mm)/sqrt(n())
  )

ggplot(sumbats2,aes(x=Species,y=body_length_mm,colour=Species))+
  geom_point()+
  geom_errorbar(aes(ymin=meanbats-sebats,ymax=meanbats+sebats))+
  theme_bw()
```

## **3. Is the number of ticks found on the bats associated with their sex or age?**

To figure this out, I'll first plot this proposed relationship, knowing that sex and age (the predictors) are categorical while tick count (the response) is a continuous variable, so I'll use a histogram, and will later run a two-way anova?

lets plot the data for now

```{r}

ggplot(bats, aes(x = tick_num, fill = sex,width=0.9)) + 
  geom_bar(position="dodge")+
  facet_wrap(~ age) + 
  theme_bw()

```

This graph is not very intuitive to read, but shows the variation across tick numbers well.

```{r}
ggplot(bats,aes(x=age,y=tick_num))+
  geom_point()+
  geom_boxplot()+
  facet_wrap(vars(sex))+
  theme_bw()
```

Here, it looks like there is a difference in mean tick num between males and females, but not between adults and subadults

next, lets calculate mean and sd for all combinations

```{r}
sumbats<-bats%>%
  group_by(sex, age)%>%
  summarise(meanbats =mean(tick_num))
sumbats
```

yes, this confirms that the mean ticks appear different between sex, but not age

Next, I'll build my statistical model. Since I'm going to run an ANOVA, I'll need to start with a linear model with an interaction

```{r}
bats.MOD3 <- lm(tick_num~sex*age, data=bats)
```

I'll need to check my assumptions before running any tests

```{r}
autoplot(bats.MOD3, smooth.colour=NA)
```

These look slightly more questionable than the previous, but nothing too alarming.A linear model is appropriate, there is normality in the data, there is even variance, and no points with major leverage.

Now I can move onto my tests

```{r}
anova(bats.MOD3)
```

The variation explained by sex is 111.063, while the variation explained by age and the interaction variable are both \<3. The sex variable has a p value of \<2e-16, suggesting we reject our hypothesis that the number of ticks found on the bats is not associated with either their sex or age. Further analysis suggests that the number of ticks found on the bats is associated with their sex, while there is no significant impact of age nor a sex/age interaction on tick number in this model (F=116.6779, df=1, p\<2e-16), (F=2.7431,df=1,p=0.1003), (F=2.4668,df=1,p=0.1189)

```{r}

sumbats3<-bats%>%
  group_by(sex, age)%>%
  summarise(
    meanbats = mean(tick_num),
    sebats=sd(tick_num)/sqrt(n())
    )
sumbats3

ggplot(sumbats3,aes(x=age,y=meanbats,colour=sex))+
  geom_point(size=2)+
  geom_errorbar(aes(ymin=meanbats-sebats,ymax=meanbats+sebats,width=0.7))+
  theme_bw()


```

## **4. Disregarding species, is there a relationship in bats between tail length and body length?**

To figure this out, I'll first plot this proposed relationship, knowing that tail length and body length are continuous variables, so I'll use a scatterplot, and will later run a linear regression model.

```{r}
ggplot(bats,aes(x=tail_length_mm,y=body_length_mm))+
  geom_point()+
  theme_bw()
```

Oooh! There appears to be a strong positive relationship. (180-60)/(100-11)=120/89 = 1.35 I estimate the slope to be \~1.35 and the y intercept to be around 30.

Next, I'll build my statistical model. Since I'm going to run an ANOVA, I'll need to start with a linear model

```{r}
bats.MOD4 <- lm(body_length_mm~tail_length_mm, data=bats)
bats.MOD4
```

So, here the slope is 1.455 and the y-int is 33.300. My guesses weren't too far off!

I'll need to check my assumptions before running any tests

```{r}
autoplot(bats.MOD4)
```

These charts look good! A linear model is appropriate, there is normality in the data, there is even variance, and no points with major leverage.

Now I can move onto my tests

```{r}
summary(bats.MOD4)
```

Woooop! We have a (very) significant p value.

I reject the null hypothesis that mean body length and tail length have no relationship, and suggest that as body length increases, so does tail length (DF1,123,F=1037, p\<2.2e-16).'

New graph

```{r}

ggplot(bats,aes(x=tail_length_mm,y=body_length_mm))+
  geom_point()+ 
  geom_smooth(method='lm')+
  theme_bw()
```
