---
title: "O'Brien-PS3"
author: "S.O'Brien"
format: html
editor: 
  markdown: 
    wrap: 72
---

### Set things up

```{r}
rm(list=ls())
library(tidyverse)
library(here)
library(ggfortify)
library(multcomp)
```

### Import data

```{r}
bat <- read.csv(here("Data/bat_data.csv"))
```

### Check out data w/ glimpse()

```{r}
glimpse(bat)
```

we see 9 different variables w/ 125 observations. Species and
common_name can be changed to a fct format bc there are only 5 levels
for each. age, sex, and type can be changed to a fct format bc they're
nominal with only 2 outcomes for them.

### Check out data w/ summary()

```{r}
summary(bat)
```

there doesn't seem to be anything out of the ordinary or the numeric
columns in the data set. there's a possibility that body_length_mm and
tail_length_mm could have an outlier.

### Change Species to fct

```{r}
bat$Species <- 
  as.factor(bat$Species)
levels(bat$Species) # shows all of the levels for Species
```

doesn't look like there are any errors

### Change common_name to fct

```{r}
bat$common_name <- 
  as.factor(bat$common_name)
levels(bat$common_name) # shows all of the levels for common_name
```

doesn't look like there are any errors

### Change age to fct

```{r}
bat$age <- 
  as.factor(bat$age)
levels(bat$age) # shows all of the levels for age
```

doesn't look like there are any errors

### Change sex to fct

```{r}
bat$sex <- 
  as.factor(bat$sex)
levels(bat$sex) # shows all of the levels for sex
```

doesn't look like there are any errors

### Change type to fct

```{r}
bat$type <- 
  as.factor(bat$type)
levels(bat$type) # shows all of the levels for type
```

doesn't look like there are any errors

### Error check mass_g variable

```{r}
# error check mass_g using a histogram
ggplot(bat, aes(mass_g)) +
  geom_histogram()

# check for missing values in mass_g
bat |>
  filter(mass_g == "NA")
```

doesn't look like there are any errors for this variable. 30g doesn't
look like an error and there arent any missing values

### Error check tick_num variable

```{r}
# error check tick_num using a histogram
ggplot(bat, aes(tick_num)) +
  geom_histogram()

# check for missing values in tick_num
bat |> 
  filter(tick_num == "NA")
```

from the histogram, there doesn't seem to be any strange values that
would raise concern. there also arent any missing values

### Error check body_length_mm

```{r}
# error check body_length_mm using hostogram
ggplot(bat, aes(body_length_mm)) +
  geom_histogram()

# check for missing values in body_length_mm
bat |>
  filter(body_length_mm == "NA")
```

there are values for body length close to 180mm but there are what seems
to be 2 counts for it so i don't think this was an error. there also
arent any missing values

### Error check tail_length_mm

```{r}
# error check tail_length_mm using hostogram
ggplot(bat, aes(tail_length_mm)) +
  geom_histogram()

# check for missing values in tail_length_mm
bat |>
  filter(tail_length_mm == "NA")
```

there are no missing values and i dont think the values on the outside
of the histogram are concerning. graph looks ok.

### Questions to be answered:

A. Does body mass differ between these five species of bats and, if so,
how does body mass differ between the species?

Response: body mass (ratio). Predictor: species (categorical). Model:
one-way ANOVA b/c response variable is continuous, predictor variable is
categorical and we're going to look at how body mass is different among
species

B. Does body length differ between these five species of bats and, if
so, how does body length differ between the species?

Response: body length (ratio). Predictor: species (categorical). Model:
one-way ANOVA b/c response variable is continuous, predictor variable is
categorical and we're going to look at how body length is different
among species

C. Is the number of ticks found on the bats associated with their sex or
age?

Response: tick number (discrete). Predictor: sex and age (categorical).
Model: chi-sq test of independence (contingency table) b/c we're looking
at how the count data (number of ticks) is associated with sex and age

D. Disregarding species, is there a relationship in bats between tail
length and body length?

Response: tail length (ratio). Predictor: body length (ratio). Model:
simple linear regression because we're comparing the relationship of two
continuous variables

# A.

Does body mass differ between these five species of bats and, if so, how
does body mass differ between the species?

Response: body mass (ratio). Predictor: species (categorical). Model:
one-way ANOVA b/c response variable is continuous, predictor variable is
categorical and we're going to look at how body mass is different among
species

### Questions:

1.  Is there a difference in mean body mass between species?

2.  If there is a difference, which species does the difference exist?

### Hypotheses

Null hypothesis: there is no difference in mean body mass between
species. Alternative hypothesis: there is a difference in mean body mass
between species.

### Plot the data

```{r}
# histograms of mass for each species
ggplot(bat, aes(mass_g)) +
  geom_histogram() +
  theme_bw() +
  facet_wrap(~ Species, ncol = 1)

# side-by-side boxplots of mass for each species
ggplot(bat, aes(Species, mass_g)) +
  geom_boxplot() +
  theme_bw()
```

There is a heavier and lighter outlier for EPFU bats and lighter outlier
for LACI bats.

I predict that there will be a significant difference in body mass for
species. My prediction is that the LACI bats will be significantly
heavier than LANO, MYLU, and PESU bats but not EPFU bats.

EPFU mean: around 21g, LACI mean: around 27g LANO mean: around 14g, MYLU
mean: around 11g, PESU mean: around 6g

### Build model

```{r}
mass_model <- lm(mass_g ~ Species, data = bat) # this anova model is called mass_model
```

### Check assumptions

```{r}
autoplot(mass_model, smooth.colour = NA)
```

Assumptions look good. Nothing is too out of the ordinary. There is some
deviation from the line on the Normal Q-Q plot.

### Interpret mass_model using anova()

```{r}
anova(mass_model)
```

We can reject the null hypothesis that there is no difference in mean
body mass between species. There is a significant effect of bat species
on the mass of bats caught (F(4,120) = 792.9, p = 2.2 x 10-16). The LACI
bats have the highest mean body mass, then EPFU, LANO, MYLU and lastly,
PESU with the lowest mean body mass.

### Interpret mass_model using summary()

```{r}
summary(mass_model)
```

Since the EPFU species is first alphabetically, this model compares all
the species against EPFU. From this, we learn that there is very strong
evidence that all of the species have significantly different body
masses from the EPFU bats. EPFU bats have a mean body mass of 20.532,
LACI bats are significantly heavier, and LANO, MYLU, and PESU bats are
all significantly lighter than EPFU bats. we can go further into this by
conducting a Tukey HSD test to analyze the difference between each
species.

### Tukey HSD test

```{r}
mass_tukey <- glht(mass_model, linfct = mcp(Species = "Tukey")) # this makes our tukey model 

summary(mass_tukey) # this shows comparisons between every species

cld(mass_tukey) # this shows compact letter display of tukey results
```

From the Tukey test, we can determine that there is very strong evidence
for a difference in mean body mass (g) between every species of bat.

### Get actual mass means for each bat species

```{r}
sumDat <- bat |> 
  group_by(Species) |>
  summarise(mean_mass = mean(mass_g))
```

I'm creating this data frame of mean weight for bat species to use in
the plot below to better visualize the data.

### Final Plot

```{r}
ggplot(bat, aes(Species, mass_g, color = Species))+
     geom_point(size = 3, alpha = 0.4) +
     geom_point(data = sumDat, aes(Species, mean_mass, fill = Species), shape = 23, size = 4)+
     coord_flip()+
     theme_bw() +
  xlab("Bat Species") +
  ylab("Mass (g)") +
  labs(caption = "Mass (g) of individual bats and mean mass for each bat species") +
  theme(plot.caption = element_text(face = "italic", size = 10, hjust = 0))
```

This plot shows the weights of different bat species with the mean
weight of each plotted as a diamond on the figure.

# B.

Does body length differ between these five species of bats and, if so,
how does body length differ between the species?

Response: body length (ratio). Predictor: species (categorical). Model:
one-way ANOVA b/c response variable is continuous, predictor variable is
categorical and we're going to look at how body length is different
among species

### Questions:

1.  Is there a difference in mean body length between species?

2.  If there is a difference, which species does the difference exist?

### Hypotheses

Null hypothesis: there is no difference in mean body length between
species. Alternative hypothesis: there is a difference in mean body
length between species.

### Plot the data

```{r}
# histograms of length for each species
ggplot(bat, aes(body_length_mm)) +
  geom_histogram() +
  theme_bw() +
  facet_wrap(~ Species, ncol = 1)

# side-by-side boxplots of length for each species
ggplot(bat, aes(Species, body_length_mm)) +
  geom_boxplot() +
  theme_bw()
```

There is a longer outlier for EPFU bats, longer and shorter outlier for
LANO bats, and shorter outlier for MYLU bats.

I think there will be a significant difference in body length for bat
species. My prediction is that the LACI bats will be significantly
longer than EPFU, LANO, MYLU, and PESU bats.

Mean body length for EPFU is around 121mm, LACI bats: around 140mm, LANO
bats: around 105mm, MYLU bats: around 95mm, PESU bats: around 85mm

### Build model

```{r}
length_model <- lm(body_length_mm ~ Species, data = bat) # this anova model is called length_model
```

### Check assumptions

```{r}
autoplot(length_model, smooth.colour = NA)
```

The assumptions look good there isn't anything too out of the ordinary.
Normality looks good except one negative value that deviates from the
line. But overall it looks okay.

### Interpret mass_model using anova()

```{r}
anova(length_model)
```

We can reject the null hypothesis that there is no difference in mean
body length between species. There is a significant effect of bat
species on the length of bats caught (F(4,120) = 70.498, p = 2.2 x
10-16). The LACI bats have the highest mean body mass, then EPFU, LANO,
MYLU and lastly, PESU with the lowest mean body mass. We can use the
summary() function to observe how mean body mass of bat species compares
to EPFU bats.

### Interpret mass_model using summary()

```{r}
summary(length_model)
```

Since the EPFU species is first alphabetically, this model compares all
the species against EPFU as the intercept. From this, we learn that
there is very strong evidence that all of the species have significantly
different body lengths from the EPFU bats. EPFU bats have a mean body
length of 123.4, LACI bats are significantly longer, and LANO, MYLU, and
PESU bats are all significantly shorter. We can go further into this by
conducting a Tukey HSD test to analyze the difference between each
species.

### Tukey HSD test

```{r}
length_tukey <- glht(length_model, linfct = mcp(Species = "Tukey")) # this makes our tukey model 

summary(length_tukey) # this shows comparisons between every species

cld(length_tukey) # this shows compact letter display of tukey results
```

From this Tukey test, we can determine that there is very strong
evidence of a difference in body length between EPFU bats and every
other species of bat. There is also very strong evidence that LACI and
LANO bats have significantly different body lengths from every other
species of bat. However, there is not evidence to support that MYLU and
PESU bats have significantly different body lengths from each other.
There is evidence that they do have significantly different body lengths
from LANO, LACI, and EPFU bat species.

### Get actual length means for each bat species

```{r}
sum_length <- bat |> 
  group_by(Species) |>
  summarise(mean_length = mean(body_length_mm))
```

I'm creating this data frame of mean length of each bat species to use
in the plot below to better visualize the data.

### Final Plot

```{r}
ggplot(bat, aes(Species, body_length_mm, color = Species))+
     geom_point(size = 3, alpha = 0.4) +
     geom_point(data = sum_length, aes(Species, mean_length, fill = Species), shape = 23, size = 4)+
     coord_flip()+
     theme_bw() +
  xlab("Bat Species") +
  ylab("Body Length (mm)") +
  labs(caption = "Body length (mm) of individual bats and mean body length for each bat species") +
  theme(plot.caption = element_text(face = "italic", 
                                    size = 10, 
                                    hjust = 0))
```

This plot shows the lengths of different bats with the mean length of
each species plotted as a diamond on the figure.

# C.

Is the number of ticks found on the bats associated with their sex or
age?

Response: tick number (discrete). Predictor: sex and age (categorical).
Model: chi-sq test of independence (contigency table) b/c we're looking
at how the count data (number of ticks) is associated with sex and age

### Question

Is there an association of sex or age of bats on the total number of
ticks found on them?

### Hypotheses

Null hypothesis: there is no association between tick number from sex or
age of bats. Alternative hypothesis: there is an association between
tick number from sex or age of bats.

### Calculate totals of tick caught across sex and age of bats

```{r}
tick_age <- bat |>
  group_by(age) |>
  summarise(ticks_collected = sum(tick_num)) # this code creates a data frame (tick_age) for the number of ticks collected (ticks_collected) for adults and subadults

tick_sex <- bat |>
  group_by(sex) |>
  summarise(ticks_collected = sum(tick_num)) # this code creates a data frame (tick_sex) for the number of ticks collected (ticks_collected) for males and females
```

For age, there were 345 ticks collected on adults and 315 ticks
collected on subadults. For sex, there were 295 ticks collected on
females and 365 ticks collected on males. Now I will plot this data as
bar charts

### Plot the data

```{r}
ggplot(tick_age, aes(x = age, y = ticks_collected)) +
  geom_bar(stat = 'identity', position = 'dodge') +
  theme_bw() # this bar char is the number of ticks on adults and subadults

ggplot(tick_sex, aes(x = sex, y = ticks_collected)) +
  geom_bar(stat = 'identity', position = 'dodge') +
  theme_bw() # this bar char is the number of ticks on males and females

# stat = 'identity' tells ggplot not to calculate anything, 'position' = 'dodge' tells ggplot to make side-by-side bar chart
# chose to do bar charts because tick_num is countable data of the number of ticks found on bats
```

From this bar chart, I predict that the number of ticks collected on
bats is not equally distributed among the sex of bats, but it is equally
distributed among the age of bats.

### Making separate age and sex matrices of data for chi-sq test

```{r}
# this is a matrix for the number of ticks found on adults and subadults
age_mat <- xtabs(tick_num ~ age,
                 data = bat) 
age_mat

# this is a matrix for the number of ticks found on males and females
sex_mat <- xtabs(tick_num ~ sex, 
                 data = bat)
sex_mat
```

Output is the same as the data frames but it's in matrix form

### Conduct chi-sq tests for sex and age

```{r}
# chi-sq test for ticks collected on adults and subadults 
chisq.test(age_mat) 

# chi-sq test for ticks collected on males and females
chisq.test(sex_mat)
```

From the chi-sq tests, we can reject the null hypothesis that there is
no association between tick number from sex or age of bats. There is a
high probability that the pattern of ticks found on males and females
didn't arise by chance and that there is an association for ticks found
and the sex of bats (X-squared = 7.4242, df = 1, p = 0.006435). However,
there is strong evidence that there is no association between the number
of ticks found and the age of bats (X-squared = 1.3636, df = 1, p =
0.2429). Overall, ticks are more frequent on male bats than female bats
but there is no evidence of ticks being more frequent on adult or
subadult bats.

### Dataframe of number of ticks for sex and age

```{r}
# this code creates a data frame (tick_sum) for the number of ticks collected (ticks_collected) across sex and age
tick_sum <- bat |>
  group_by(sex, age) |>
  summarise(ticks_collected = sum(tick_num)) 
```

This data frame will be used in the plot below.

### Final plot

```{r}
ggplot(tick_sum, aes(x = sex, y = ticks_collected, fill = age)) +
  geom_bar(stat = 'identity', position = 'dodge') +
  theme_bw() +
  xlab("Sex of Bat") +
  ylab("Number of Ticks Collected") +
  labs(caption = "Number of ticks collected across sex and age of bats.") +
  theme(plot.caption = element_text(face = "italic", 
                                    size = 10, 
                                    hjust = 0)) +
  scale_fill_discrete(name = "Age")
```

This plot shows the total number of ticks collected for each sex and age
of bat. We can see that there isn't much difference for age, especially
for females, but there is more of a difference for sex.

# D.

Disregarding species, is there a relationship in bats between tail
length and body length?

Response: tail length (ratio). Predictor: body length (ratio). Model:
simple linear regression because we're comparing the relationship of two
continuous variables

### Question

Is there a relationship between tail length and body length of bats?

### Hypotheses

Null hypothesis: there is no relationship between tail length and body
length of bats. Alternative hypothesis: there is a relationship between
tail length and body length of bats.

### Plot the data

```{r}
ggplot(bat, aes(x = body_length_mm, y = tail_length_mm)) +
  geom_point() +
  theme_bw() +
  ylab("Tail Length") +
  xlab("Body Length")
```

The slope is positive so the longer bodies have longer tails. The body
length varies between 60-180mm and the tail length varies between
10-100mm. therefore, I estimate the slope to be around 90/120 = 0.75. I
would guess the y-intercept to be around 10. Df for error = 125 - 2 =
123.

### Make simple linear regression model

```{r}
body_tail_model <- lm(tail_length_mm ~ body_length_mm,
                      data = bat)
```

Let's check assumptions first.

### Check assumptions

```{r}
autoplot(body_tail_model, smooth.colour = NA)
```

Residuals-fitted: looks equally distributed, good. no humps or valleys.

Normal Q-Q: dots all fall pretty close to the expected values line,
good.

Scale-location: no pattern, all spread randomly.

Residuals-leverage: more dots condensed to the left with some scattered
out right

smooth.colour = NA suppresses squiggly line fitted by the regression.

### Interpret body_tail_model using anova() and summary()

```{r}
anova(body_tail_model)

summary(body_tail_model)
```

From these models, we can reject the null hypothesis that there is no
relationship between body length and tail length. There is very strong
evidence that body and tail length are related, as one increases then
the other increases as well (adj R2 = 0.8931, p = 2.2 x 10-16).

### Final plot

```{r}
ggplot(bat, aes(x = body_length_mm, y = tail_length_mm)) +
  geom_point() +
  geom_smooth(method = "lm") + # geom_smooth adds fitted values and standard error of fit to the figure
  labs(x = "Body Length (mm)", 
       y = "Tail Length (mm)", 
       caption = "Scatterplot of body length (mm) and tail length (mm) with the fitted values and the standard error of the fit added.") +
  theme(plot.caption = element_text(face = "italic", 
                                    size = 10, 
                                    hjust = 0)) +
  theme_bw()
```

This plot is the same scatter plot as the previous figure but 'method =
"lm"' in geom_smooth() adds fitted values and standard error of fit to
the figure. As the lengths increase past 130 and decrease past 80 then
the standard error increases.

# This concludes my analysis, hope it helps!!
